---
title: 'Assignment 2: Coding Basics'
author: "Shaochong Xue"
output:
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

```{r, include=FALSE}
rm(list = ls())
```


## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. Generating a sequence from 0 to 55, increasing by 5.
seq55 <- seq(0, 55, by = 5) #if the sequence starts at 1 and increases by 5 , it won't reach 55
seq55

#2. Calculating the mean and median of the sequence.
mean_seq <- mean(seq55)
mean_seq

median_seq <- median(seq55)
median_seq

#3. Comparing the mean and median.
mean_seq > median_seq
mean_seq == median_seq

```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
student_names <- c("A", "B", "C", "D")                # Vector type: character
test_scores <- c(95, 90, 75, 85)                      # Vector type: numeric
scholarship <- c(TRUE, FALSE, TRUE, FALSE)         # Vector type: logical

student.df <- data.frame(
  Name = student_names, 
  Score = test_scores, 
  Scholarship = scholarship
  )
student.df
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:  A data frame can have different data types and labled data, while a matrix can only have one data type for all elements. 

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail". 

```{r function}

function1 <- function(score) {
  if (score > 50) {
    print("Pass")
  } else {
    print("Fail") 
  }
}

function1(100)  
function1(20)   

```


11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

```{r function2}

function2 <- function(score) {
  result <- ifelse(score > 50, "Pass", "Fail") 
  print(result) 
}

function2(100)  
function2(20)  

```


12. Run both functions using the value 52.5 as the input
```{r}
function1(52.5)
function2(52.5)
```


13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#10. Create a function using if...else

function1 <- function(score) {
  if (score > 50) {
    print("Pass")
  } else {
    print("Fail") 
  }
}

function1(100)  
function1(20)   

#11. Create a function using ifelse()

function2 <- function(score) {
  result <- ifelse(score > 50, "Pass", "Fail") 
  print(result) 
}

function2(100)  
function2(20)  


#12a. Run the first function with the value 52.5
function1(52.5)

#12b. Run the second function with the value 52.5
function2(52.5)
```

#13a. Run the first function with the vector of test scores
function1(student.df$Score)

```{r}
#13b. Run the second function with the vector of test scores
function2(student.df$Score)
```


14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer:`ifelse`worked correctly, because it can handle vector while `if`...`else` cannot. (giving the error:"the condition has length > 1") 


**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)