---
title: "Assignment 8: Time Series Analysis"
author: "Shaochong Xue"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r, include=FALSE}
#cleaning the environment
rm(list = ls())
```

```{r}

library(tidyverse)
library(lubridate)
#install.packages("trend")
library(trend)
#install.packages("zoo")
library(zoo)
#install.packages("Kendall")
library(Kendall)
#install.packages("tseries")
library(tseries)
library(here)

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

here()
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1

GaringerOzone <- list.files("Data/Raw/Ozone_TimeSeries", pattern = "*.csv", full.names = TRUE) %>%
  map_dfr(read.csv, stringsAsFactors = FALSE)


```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}

#3
GaringerOzone$Date <- as.Date(as.character(GaringerOzone$Date), format = "%m/%d/%Y")

#4
print(colnames(GaringerOzone))

GaringerOzone_processed <- 
  GaringerOzone %>%
  select(Date, Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE) #This chunk was able to run but failed to knit. 


# 5
Days <- tibble(Date = seq(from = as.Date("2010-01-01"), 
                          to = as.Date("2019-12-31"), 
                          by = "day"))
  
# 6
GaringerOzone_processed <- left_join(Days, GaringerOzone_processed, by = "Date")
head(GaringerOzone_processed)

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7
ozone_by_time <-
  ggplot(GaringerOzone_processed, aes(x = Date, y = Daily.Max.8.hour.Ozone.Concentration)) +
  geom_line() +
  geom_smooth(method = "lm") +
  labs(title = "Daily Maximum 8-Hour Ozone Concentration (2010-2019)",
       x = "Year",
       y = "Ozone Concentration") +
  mytheme

print(ozone_by_time)
```

>Answer: Yes, the plot indicates a decreasing trend in ozone concentration from 2010 to 2019.

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8
GaringerOzone_clean <- 
  GaringerOzone_processed %>%
  mutate(
    Daily.Max.8.hour.Ozone.Concentration = na.approx(Daily.Max.8.hour.Ozone.Concentration, na.rm = FALSE), #filling NA
    DAILY_AQI_VALUE = na.approx(DAILY_AQI_VALUE, na.rm = FALSE) 
  ) %>%
  rename(
    Ozone = Daily.Max.8.hour.Ozone.Concentration,
    AQI = DAILY_AQI_VALUE 
  )

sum(is.na(GaringerOzone_clean$Ozone)) #testing NA
sum(is.na(GaringerOzone_clean$AQI))

```

> Answer: Since the data visualization suggests possible periodic fluctuations but does not follow a strict curve function, piecewise constant interpolation is not suitable because it simply carries forward the previous value, which may not accurately reflect the periodic trend. Spline interpolation could lead to over-smoothing, distorting the original pattern, which consists of linear segments rather than smooth curves. Therefore, linear interpolation is the most appropriate choice.


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9
# Creating year and month columns
GaringerOzone.monthly <- 
  GaringerOzone_clean %>%
  mutate(
    year = year(Date),
    month = month(Date)
  ) %>%
  group_by(year, month) %>%
  summarise(
    mean_Ozone = mean(Ozone, na.rm = TRUE),
    .groups = "drop"
  )

# creating month-year
GaringerOzone.monthly <- GaringerOzone.monthly %>%
  mutate(Date = make_date(year, month)) 

```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10
# daily
GaringerOzone.daily.ts <- ts(GaringerOzone_clean$Ozone, start = c(2010,1,1), frequency = 365)

plot(GaringerOzone.daily.ts)

# monthly
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$mean_Ozone, start = c(2010,1), frequency = 12)

plot(GaringerOzone.monthly.ts)
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11
# daily
daily.decomp <- stl(GaringerOzone.daily.ts,s.window = "periodic")
plot(daily.decomp)

#monthly
monthly.decomp <- stl(GaringerOzone.monthly.ts,s.window = "periodic")
plot(monthly.decomp)

```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12
Ozone_trend <- smk.test(GaringerOzone.monthly.ts)
Ozone_trend

summary(Ozone_trend)
```

> Answer: 

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
# 13
mean_monthly_ozone <- 
  ggplot(GaringerOzone.monthly, aes(x = Date, y = mean_Ozone)) +
  geom_line(color = "steelblue", size=1) +
  geom_point(color = "red",size=2, alpha = 0.6) +
  labs(
    title = "Mean Monthly Ozone Concentrations (2010-2019)",
    x = "Year",
    y = "Mean Ozone Concentration (ppm)"
  ) +
  scale_x_date(date_breaks = "1 years", date_labels = "%Y") +
  mytheme

print(mean_monthly_ozone)
```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: The mean monthly ozone concentration plot shows seasonal fluctuations, and overall there is a slight decreasing trend over time. The Seasonal Mann-Kendall test confirms this (z = -1.963, p = 0.04965), indicating a statistically significant decline in ozone levels. The overall decrease suggests that long-term factors, such as emission reductions or atmospheric changes, may be contributing to lower ozone concentrations at this station.

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15

# subtract the seasomal component
GaringerOzone.monthly.nonseasonal <- GaringerOzone.monthly.ts - monthly.decomp$time.series[, "seasonal"]
plot(GaringerOzone.monthly.nonseasonal)



#16
Ozone_trend_noseasonal <- mk.test(GaringerOzone.monthly.nonseasonal)
print(Ozone_trend_noseasonal)

```

> Answer: The Mann-Kendall test on the non-seasonal ozone series shows a stronger and more significant decreasing trend (z = -2.672, p = 0.00754) compared to the Seasonal test. This suggests that while seasonal effects contribute to ozone variability, the overall long-term decline in ozone concentrations is significant even after removing seasonality.
