---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Shaochong Xue"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

```{r, include=FALSE}
#cleaning the environment
rm(list = ls())
```


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r libraries, include=FALSE}
library(tidyverse)
library(agricolae)   
library(here)
library(dplyr)
library(corrplot)
```


```{r setup2}
#1
chem_phys_raw <- read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
                          stringsAsFactors = TRUE)

chem_phys_raw$sampledate <- as.Date(chem_phys_raw$sampledate, format = "%m/%d/%y")

head(chem_phys_raw)
```

```{r ggplot theme}
#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
        legend.position = "top")

theme_set(mytheme)

```


## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature recorded during July does not change with depth across all lakes.
Ha: The mean lake temperature recorded during July changes with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
jul_data <- chem_phys_raw %>%
  filter(format(sampledate, "%m") == "07") %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na() 

head(jul_data)

#5
ggplot(jul_data, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.2) +  
  geom_smooth(method = "lm", color = "darkred") + 
  labs(title = "Lake Temperature vs. Depth (July)",
       x = "Depth (m)",
       y = "Temperature (°C)") +
  ylim(0, 35) + 
  mytheme

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The distribution of points suggests that the relationship is not strictly linear, as the rate of temperature decrease varies across depth ranges. In the 0-3m range, the temperature decreases slowly with depth. Between 3-8m, the decline in temperature becomes steeper. Beyond 8m, the rate of temperature decrease slows down again. This pattern indicates possible thermal stratification in the lake.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
temp_vs_depth <- lm(temperature_C ~ depth, data = jul_data)

# Display summary of regression results
summary(temp_vs_depth)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: According to the linear regression model, p value < 0.05, we reject H0, accept Ha, indicating that the mean lake temprature during July is significantly correlated to depth across all lakes. Every  1m increase in depth is associated with 1.95°C decrease in temprature. The R² is 0.7387, meaning that 73.9% of the variability in temperature is explained by depth. The df of this model is 9726. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
AIC_test <- lm(temperature_C ~ depth + year4 + daynum, data = jul_data)
step(AIC_test)

#10
temp_glm <- lm(temperature_C ~ depth + daynum, data = jul_data)
summary(temp_glm)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC suggests using depth and daynum as the explanatory variables to predict temperature. The multiple linear model explains 74.1% of the observed variance (R² = 0.741), which is slightly higher than the former model. However, we still need to check the residual distribution to determine whether this additional variable should be included.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

# ANOVA
lake_anova <- aov(temperature_C ~ lakename, data = jul_data)
summary(lake_anova)

# Linear model
lake_lm <- lm(temperature_C ~ lakename, data = jul_data)
summary(lake_lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Both ANOVA and the linear model show that there is a significant difference in mean temperature among lakes (p < 0.05). However, the R² of the linear model is relatively low (0.0395), meaning lake identity explains only a small portion of the temperature variation.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(jul_data, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", size = 0.5) +  
  labs(title = "Lake Temperature vs. Depth (July)",
       x = "Depth (m)",
       y = "Temperature (°C)",
       color = "Lake") + 
  ylim(0, 35) + 
  mytheme

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
# Tukey's HSD test
tukey_result <- TukeyHSD(lake_anova)

# Display results
print(tukey_result)

```


16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake and West Long Lake have the same mean temperature as Peter Lake (p > 0.05). East Long Lake has a mean temperature that is statistically distinct from all other lakes (p < 0.05 in all comparisons).


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A Welch’s two-sample t-test would be an appropriate method to compare the mean temperatures of Peter Lake and Paul Lake. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r filter}
crampton_vs_ward <- jul_data %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

head(crampton_vs_ward)
```

```{r normality test}
shapiro.test(crampton_vs_ward$temperature_C[crampton_vs_ward$lakename == "Crampton Lake"])
shapiro.test(crampton_vs_ward$temperature_C[crampton_vs_ward$lakename == "Ward Lake"])
```


```{r t.test}
wilcox.test(temperature_C ~ lakename, data = crampton_vs_ward)
```

>Answer: The Wilcoxon test shows no significant difference in temperature between Crampton Lake and Ward Lake (p = 0.8312). This means their mean temperatures are statistically similar. This result aligns with Tukey's HSD test from part 16, which also did not indicate a significant difference between these two lakes.
