---
title: "Assignment 10: Data Scraping"
author: "Shaochong Xue"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r, include=FALSE}
#cleaning the environment
rm(list = ls())
```


```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(lubridate)
library(here);
library(rvest)
library(ggplot2)
library(purrr)

here()

# Set theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".


```{r scrape.the.data}
#3 
water_system_name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water_system_name

PWSID <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

Ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Ownership

MGD <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text() %>%
  as.numeric()
MGD


```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

```{r}
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
year <- 2024

#create date column
date_col <- as.Date(paste(year, months, "01", sep = "-"), format = "%Y-%b-%d")

LWSP.df <- data.frame(
  Water_System = rep(water_system_name, length(months)),
  PWSID = rep(PWSID, length(months)),
  Ownership = rep(Ownership, length(months)),
  MGD = MGD,
  Month = months,
  Date = date_col
)
head(LWSP.df)

```


 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#5 
ggplot(LWSP.df, aes(x = Date, y = MGD)) +
  geom_line(color = "red") +
  labs(
    title = "Maximum Daily Water Use in 2024",
    x = "Month",
    y = "Maximum Day Use"
  ) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  mytheme

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two input - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape.LWSP <- function(PWSID, year) {
  the_website <- read_html(paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", 
                PWSID, "&year=", year))
  
  water_system_name_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
  ownership_tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
  mgd_tag <- "th~ td+ td"
  

  water_system_name <- the_website %>% html_nodes(water_system_name_tag) %>% html_text()
  Ownership <- the_website %>% html_nodes(ownership_tag) %>% html_text()
  MGD <- the_website %>% html_nodes(mgd_tag) %>% html_text() %>% as.numeric()
  
  
  df <- data.frame(
    "Month" = rep(1:12),
    "Year" = rep(year, 12),
    "MGD" = MGD
  ) %>%
    mutate(
      Water_System = water_system_name,
      PWSID = PWSID,
      Ownership = Ownership,
      Date = as.Date(paste(Year, Month, "01", sep = "-"))
    )
  

  return(df)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020

```{r fetch.and.plot.Durham.2020.data}
#7 
durham_2020 <- scrape.LWSP("03-32-010", 2020)
head(durham_2020)

```

8. Use the function above to extract data for Asheville (PWSID = '01-11-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2020.data}
#8 
asheville_2020 <- scrape.LWSP("01-11-010", 2020)
head(asheville_2020)


```

```{r}
Combined_2020 <- rbind(durham_2020, asheville_2020)
Combined_2020

ggplot(Combined_2020, aes(x = Date, y = MGD, color = Water_System, group = Water_System)) +
  geom_line() +
  labs(
    title = "Durham vs. Asheville Maximum Daily Water Use in 2020",
    x = "Month",
    y = "Maximum Day Use (MGD)",
    color = "Water System"
  ) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  mytheme
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
the_years <- 2018:2023
pwsids <- rep("01-11-010", length(the_years))

asheville_2018_2023 <- map2(pwsids, the_years, scrape.LWSP) %>%
  bind_rows()
asheville_2018_2023

ggplot(asheville_2018_2023, aes(x = Date, y = MGD)) +
  geom_line() +
  geom_smooth(method = "loess", color = "red") +
  labs(
    title = "Asheville Max Daily Water Use (2018–2023)",
    x = "Month",
    y = "Max Daily Use"
  ) +
  mytheme
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
> Yes. Asheville shows an increasing trend in maximum daily water use from 2018 to 2023. However, there appears to be a slight dip between 2020 and 2022, which might be related to the impact of COVID-19.
