---
title: 'Assignment 3: Data Exploration'
author: "Shaochong Xue"
date: "Fall 2024"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r, include=FALSE}
#cleaning the environment
rm(list = ls())
```

```{r library, include=FALSE}
#loading necessary packages
library(tidyverse)  
library(lubridate)
library(here)

```

```{r here, include=FALSE}
#setting the correct directory
getwd()
here()
```

```{r read csv, echo=FALSE}
# Importing the ECOTOX 
Neonics <- read.csv(
  file = here("Data", "Raw", "ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE
)

# Importing the NEON
Litter <- read.csv(
  file = here("Data", "Raw", "NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE
)

Neonics
Litter
```

```{r str, echo=FALSE}
# View structure 
str(Neonics)
str(Litter)
```


## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: 
Neonicotinoids are widely used insecticides in agriculture, but they can have harmful effects on insects, including bees and other curcial pollinators. Studying the ecotoxicology of neonicotinoids can helps us understand how its chemicals impact insect and the ecosystem. Since insects play important roles in pollination, pest control, and the food chain, their decline may have serious consequences for agriculture and the environment.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:
Studying litter and woody debris in forests is important because they play a key role in nutrient cycling and soil health. When leaves, branches, and wood decompose, they release essential nutrients back into the soil, supporting plant growth. Woody debris can also provides habitat for insects, salamanders, fungi, etc.. Moreover, understanding litter decomposition rates helps scientists predict carbon storage and climate change impacts in forest ecosystems.
Wildfires can also play a role in this process by rapidly breaking down organic matter and releasing nutrients into the soil.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Samples are categorized into functional groups like leaves, needles, seeds, flowers and woody material.
 2.Each sample is dried to improve the accuracy of measuring carbon content and biomass.
 3.Using sample control to ensure the accuracy of the measurement.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r dim, echo=FALSE}
# Checking the dimensions
nrow(Neonics)
length(Neonics)

```
> Answer: The dataset has 4,623 rows and 30 columns.

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r sum, echo=FALSE}
# summary 
effect_summary <- summary(Neonics$Effect)
effect_summary

```

```{r sort, echo=FALSE}
# Sorting the effects
sort_effects <- sort(effect_summary, decreasing = TRUE)
sort_effects

```


> Answer:
The most common effects studied are Population decline (1803), Mortality (1493), and Behavioral changes (360). These is important because population decline and mortality directly affect insect biodiversity, while behavioral changes, such as feeding behavior, can disrupt pollination and food web interactions. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r sum species, echo=FALSE}
# summary species
species_summary <- summary(Neonics$Species.Common.Name, maxsum = 6)

# sort
top6_species <- sort(species_summary, decreasing = TRUE)
top6_species

```

> Answer: These six most commnon species are primarily bees or bumble bees, which are pollinators or beneficial insects. They are important for agriculture and ecosystem health. Their sensitivity to neonicotinoids is of interest because declines in pollinator populations can reduce crop yields and disrupt food webs.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r numeric, echo=FALSE}
# checking the column
class(Neonics$Conc.1..Author.)

# unique values
unique(Neonics$Conc.1..Author.)

```

> Answer: The Conc.1..Author. column is a factor instead of numeric because it contains non-numeric characters like "NR", <, >, and values with /. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r polugon plot, echo=FALSE}
# polygon plot by publication year
plot1 <-
  ggplot(Neonics, aes(x = Publication.Year)) +
  geom_freqpoly(binwidth = 1) +
  scale_x_continuous(breaks = seq(min(Neonics$Publication.Year, na.rm = TRUE), 
                                  max(Neonics$Publication.Year, na.rm = TRUE), 
                                  by = 5)) +  
  labs(title = "Number of Studies by Publication Year",
       x = "Publication Year",
       y = "Number of Studies") +
  theme_minimal()

plot1
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r olugon plot 2, echo=FALSE}
# polygon plot with Test.Location
plot2 <-
  ggplot(Neonics, aes(x = Publication.Year, color = Test.Location)) +
  geom_freqpoly(binwidth = 1) +  
  scale_x_continuous(breaks = seq(min(Neonics$Publication.Year, na.rm = TRUE), 
                                  max(Neonics$Publication.Year, na.rm = TRUE), 
                                  by = 5)) +  
  labs(title = "Number of Studies by Publication Year and Test Location",
       x = "Publication Year",
       y = "Number of Studies",
       color = "Test Location") 
  theme_minimal()

plot2
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:The most common test locations is Lab. Lab studies showed an overall increasing trend, peaking around 2015, then rapidly declining. Field natural studies had a sudden increase around 2010 but remained relatively steady overall. Field artificial and Field undeterminable were much less common.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r bar graph, echo=FALSE}
# bar graph of Endpoint
plot3 <- 
  ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar(fill="steelblue") +
  labs(title = "Count of Different Endpoints in the Dataset",
       x = "Endpoint",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

plot3
```

> Answer: The two most common endpoints are LOEC and NOEC. LOEC represents the lowest concentration at which a significant effect is observed, while NOEC is the highest concentration where no significant effect is detected.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r date, echo=FALSE}
# checking the class
class(Litter$collectDate)

# converting to Date format
Litter$collectDate <- ymd(Litter$collectDate)
class(Litter$collectDate)

```

```{r aug18, echo=FALSE}
# unique aug2018
aug2018 <- unique(Litter$collectDate[month(Litter$collectDate) == 8 & year(Litter$collectDate) == 2018])

aug2018
```


13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r plot, echo=FALSE}
# unique plots at Niwot Ridge
unique_plots <- unique(Litter$plotID)
unique_plots

#plots number
length(unique_plots)

# Compare with summary()
summary(Litter$plotID)

```

> Answer:  
There are 12 different plots sampled at Niwot Ridge.
unique() provides a list of all distinct plot IDs, while summary() shows how many times each plot was sampled.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r bar graph 2, echo=FALSE}
# bar graph of functionalGroup counts
plot4 <-
  ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar(fill = "orange") + 
  labs(title = "Count of Functional Groups in Niwot Ridge",
       x = "Functional Group",
       y = "Count") +
  theme_minimal() 

plot4
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r boxplot, echo=FALSE}
# creating boxplot
plot5 <-
  ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_boxplot(fill = "lightblue") +  
  labs(title = "Boxplot of Dry Mass by Functional Group",
       x = "Functional Group",
       y = "Dry Mass") +
  theme_minimal()

plot5
```

```{r violin plot, echo=FALSE}
# create violin plot
plot6 <-
  ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_violin(fill = "violet",bw = 0.5, draw_quantiles = c(0.25, 0.5, 0.75)) +
  labs(title = "Violin Plot of Dry Mass by Functional Group",
       x = "Functional Group",
       y = "Dry Mass") +
  theme_minimal()
plot6
```


Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:
The boxplot is more effective in this case because the dryMass data is highly skewed with many small values.It's hard for violin plot to represent these distributions clearly, while boxplot effectively shows medians, quartiles, and outliers, making it easier to compare different functional groups and identify extreme values.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
Needles has the highest biomass at these sites, with largest range and highest values overall. 
